Taking uncertainty in word embedding bias estimation seriously: a bayesian approach

### 2022-06-22 paperDraft/paperDraft4.tex

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  10pt,
  dvipsnames,enabledeprecatedfontcommands, twocolumn]{scrartcl}
\title{Taking uncertainty in word embedding bias estimation seriously: a
bayesian approach}
\author{Alicja Dobrzeniecka and Rafal Urbaniak}
\date{}

\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Taking uncertainty in word embedding bias estimation seriously: a bayesian approach},
  pdfauthor={Alicja Dobrzeniecka and Rafal Urbaniak},
  colorlinks=true,
  linkcolor={Maroon},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={blue},
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
%\documentclass{article}

@@

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

@@

\raf{fix bibliography}

\ali{proofread and think if we should add more}

@@

@@

The effect size is computed by normalizing the difference in means as
follows:\footnote{ WEAT is a modification of the Implicit Association
  Test (IAT) {[}8{]} used in psychology, and it uses almost the same
  word sets, allowing for a \emph{prima facie} sensible comparison with
  bias in humans. {[}2{]} argue that significant biases---thus
  measured--- similar to the ones discovered by IAT can be discovered in
  word embeddings. {[}5{]} extended the methodology to a multilingual
  and cross-lingual setting, arguing that using Euclidean distance
  instead of similarity does not make much difference, while the bias
  effects vary greatly across embedding models (interestingly, with
  social media-text trained embeddings being less biased than those
  based on Wikipedia).}\(^{, }\)\footnote{ A similar methodology is
  employed in {[}3{]}. The authors employ word embeddings trained on
  corpora from different decades to study the shifts in various biases.
  For instance, to compute the occupational embeddings bias for women
  the authors first compute the average vector of vector embeddings of
  words that represent women (e.g.~\emph{she}, and \emph{female}), then
  calculate the Euclidean distance between this mean vector and words
  for occupations. Then they take the mean of these distances and
  subtract from it the analogously obtained mean for the average vector
  of vector embeddings of words that represent men. Formally they take
  the relative norm distance between \(X\) and \(Y\) to be:}

@@

\hypertarget{non-statistical-problems-with-cosine-based-measures-of-bias}{%
\section{Non-statistical problems with cosine-based measures of
bias}\label{non-statistical-problems-with-cosine-based-measures-of-bias}}

@@

\hypertarget{statistical-issues}{%
\section{Statistical issues}\label{statistical-issues}}

@@

\hypertarget{refs}{}
\begin{CSLReferences}{0}{0}
\leavevmode\vadjust pre{\hypertarget{ref-Bolukbasi2016man}{}}%
\CSLLeftMargin{{[}1{]} }
\CSLRightInline{Tolga Bolukbasi, Kai-Wei Chang, James Y. Zou, Venkatesh
Saligrama, and Adam Kalai. 2016. Man is to computer programmer as woman
is to homemaker? Debiasing word embeddings. \emph{CoRR} abs/1607.06520,
(2016). Retrieved from \url{http://arxiv.org/abs/1607.06520}}

\leavevmode\vadjust pre{\hypertarget{ref-Caliskan2017semanticsBiases}{}}%
\CSLLeftMargin{{[}2{]} }
\CSLRightInline{Aylin Caliskan, Joanna J. Bryson, and Arvind Narayanan.
2017. Semantics derived automatically from language corpora contain
human-like biases. \emph{Science} 356, 6334 (April 2017), 183--186.
DOI:https://doi.org/\href{https://doi.org/10.1126/science.aal4230}{10.1126/science.aal4230}}

\leavevmode\vadjust pre{\hypertarget{ref-Garg2018years}{}}%
\CSLLeftMargin{{[}3{]} }
\CSLRightInline{Nikhil Garg, Londa Schiebinger, Dan Jurafsky, and James
Zou. 2018. Word embeddings quantify 100 years of gender and ethnic
stereotypes. \emph{Proceedings of the National Academy of Sciences} 115,
16 (April 2018), E3635--E3644.
DOI:https://doi.org/\href{https://doi.org/10.1073/pnas.1720347115}{10.1073/pnas.1720347115}}

\leavevmode\vadjust pre{\hypertarget{ref-Gonen2019lipstick}{}}%
\CSLLeftMargin{{[}4{]} }
\CSLRightInline{Hila Gonen and Yoav Goldberg. 2019. Lipstick on a pig:
{D}ebiasing methods cover up systematic gender biases in word embeddings
but do not remove them. In \emph{Proceedings of the 2019 conference of
the north {A}merican chapter of the association for computational
linguistics: Human language technologies, volume 1 (long and short
papers)}, Association for Computational Linguistics, Minneapolis,
Minnesota, 609--614.
DOI:https://doi.org/\href{https://doi.org/10.18653/v1/N19-1061}{10.18653/v1/N19-1061}}

\leavevmode\vadjust pre{\hypertarget{ref-Lauscher2019multidimensional}{}}%
\CSLLeftMargin{{[}5{]} }
\CSLRightInline{Anne Lauscher and Goran Glavas. 2019. Are we
consistently biased? Multidimensional analysis of biases in
distributional word vectors. \emph{CoRR} abs/1904.11783, (2019).
Retrieved from \url{http://arxiv.org/abs/1904.11783}}

\leavevmode\vadjust pre{\hypertarget{ref-Manzini2019blackToCriminal}{}}%
\CSLLeftMargin{{[}6{]} }
\CSLRightInline{Thomas Manzini, Yao Chong Lim, Yulia Tsvetkov, and Alan
W Black. 2019. Black is to criminal as caucasian is to police: Detecting
and removing multiclass bias in word embeddings. Retrieved from
\url{https://arxiv.org/abs/1904.04047}}

\leavevmode\vadjust pre{\hypertarget{ref-Nissim2020fair}{}}%
\CSLLeftMargin{{[}7{]} }
\CSLRightInline{Malvina Nissim, Rik van Noord, and Rob van der Goot.
2020. Fair is better than sensational: Man is to doctor as woman is to
doctor. \emph{Computational Linguistics} 46, 2 (June 2020), 487--497.
DOI:https://doi.org/\href{https://doi.org/10.1162/coli_a_00379}{10.1162/coli\_a\_00379}}

\leavevmode\vadjust pre{\hypertarget{ref-Nosek2002harvesting}{}}%
\CSLLeftMargin{{[}8{]} }
\CSLRightInline{Brian A. Nosek, Mahzarin R. Banaji, and Anthony G.
Greenwald. 2002. Harvesting implicit group attitudes and beliefs from a
demonstration web site. \emph{Group Dynamics: Theory, Research, and
Practice} 6, 1 (2002), 101--115.
DOI:https://doi.org/\href{https://doi.org/10.1037/1089-2699.6.1.101}{10.1037/1089-2699.6.1.101}}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames*,x11names*}{xcolor}
%
\documentclass[
  10pt,
  dvipsnames,enabledeprecatedfontcommands, twocolumn]{scrartcl}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Taking uncertainty in word embedding bias estimation seriously: a bayesian approach},
  pdfauthor={Alicja Dobrzeniecka and Rafal Urbaniak},
  colorlinks=true,
  linkcolor=Maroon,
  filecolor=Maroon,
  citecolor=Blue,
  urlcolor=blue,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
%\documentclass{article}

@@

\ifluatex
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\title{Taking uncertainty in word embedding bias estimation seriously: a
bayesian approach}
\author{Alicja Dobrzeniecka and Rafal Urbaniak}
\date{}

@@

\hypertarget{what-is-bias}{%
\subsection{What is bias?}\label{what-is-bias}}

\hypertarget{word-embeddings}{%
\subsection{Word embeddings}\label{word-embeddings}}

@@

\hypertarget{semantical-interpretation}{%
\subsection{Semantical interpretation}\label{semantical-interpretation}}

\hypertarget{general-challenges}{%
\subsection{General challenges}\label{general-challenges}}

\hypertarget{weat-and-mac-measures}{%
\subsection{WEAT and MAC measures}\label{weat-and-mac-measures}}

@@

The effect size is computed by normalizing the difference in means as
follows:\footnote{ WEAT is a modification of the Implicit Association
  Test (IAT) {[}8{]} used in psychology, and it uses almost the same
  word sets, allowing for a \emph{prima facie} sensible comparison with
  bias in humans. {[}2{]} argue that significant biases---thus
  measured--- similar to the ones discovered by IAT can be discovered in
  word embeddings. {[}5{]} extended the methodology to a multilingual
  and cross-lingual setting, arguing that using Euclidean distance
  instead of similarity does not make much difference, while the bias
  effects vary greatly across embedding models (interestingly, with
  social media-text trained embeddings being less biased than those
  based on Wikipedia).}\(^{, }\) \^{}{[} A similar methodology is
employed in {[}3{]}. The authors employ word embeddings trained on
corpora from different decades to study the shifts in various biases.
For instance, to compute the occupational embeddings bias for women the
authors first compute the average vector of vector embeddings of words
that represent women (e.g.~\emph{she}, and \emph{female}), then
calculate the Euclidean distance between this mean vector and words for
occupations. Then they take the mean of these distances and subtract
from it the analogously obtained mean for the average vector of vector
embeddings of words that represent men. Formally they take the relative
norm distance between \(X\) and \(Y\) to be:

\begin{align}
\textsf{relative norm distance} & = \sum_{v_m\in M} \vert \vert v_m - v_X\vert \vert_2 - \vert v_m - v_Y\vert \vert_2
\end{align}

\noindent where the norm used is Euclidean, and \(v_X\) and \(x_Y\) are
average vectors for sets \(X\) and \(Y\) respectively. {]}

@@

\hypertarget{non-statistical-problems-with-cosine-based-measures-of-bias}{%
\subsection{Non-statistical problems with cosine-based measures of
bias}\label{non-statistical-problems-with-cosine-based-measures-of-bias}}

@@

\hypertarget{statistical-issues}{%
\subsection{Statistical issues}\label{statistical-issues}}

@@

\hypertarget{pre-averaging-and-manufactured-certainty}{%
\section{Pre-averaging and manufactured
certainty}\label{pre-averaging-and-manufactured-certainty}}

\hypertarget{general-problem-with-pre-averaging}{%
\subsection{General problem with
pre-averaging}\label{general-problem-with-pre-averaging}}

\hypertarget{simulations-for-weat}{%
\subsection{Simulations for WEAT}\label{simulations-for-weat}}

\hypertarget{simulations-for-mac}{%
\subsection{Simulations for MAC}\label{simulations-for-mac}}

\hypertarget{bayesian-method}{%
\section{Bayesian method}\label{bayesian-method}}

\hypertarget{bernstein-approach}{%
\subsection{Bernstein approach}\label{bernstein-approach}}

\hypertarget{chasing-metrics}{%
\subsection{Chasing metrics}\label{chasing-metrics}}

\hypertarget{bayesian-method-introduction}{%
\subsection{Bayesian method
introduction}\label{bayesian-method-introduction}}

\hypertarget{existing-applications-to-nlp-and-perhaps-to-bias}{%
\subsection{Existing applications to NLP and perhaps to
bias}\label{existing-applications-to-nlp-and-perhaps-to-bias}}

\hypertarget{model}{%
\subsection{Model}\label{model}}

\hypertarget{results}{%
\section{Results}\label{results}}

\hypertarget{weat}{%
\subsection{WEAT}\label{weat}}

\hypertarget{mac}{%
\subsection{MAC}\label{mac}}

\hypertarget{discussion-and-summary}{%
\section{Discussion and summary}\label{discussion-and-summary}}

\hypertarget{appendix}{%
\section{APPENDIX}\label{appendix}}

\hypertarget{examples-of-weat-and-mac-calculations}{%
\subsection{Examples of WEAT and MAC
calculations}\label{examples-of-weat-and-mac-calculations}}

\hypertarget{word-lists}{%
\subsection*{Word lists}\label{word-lists}}
\addcontentsline{toc}{subsection}{Word lists}

\hypertarget{refs}{}
\begin{CSLReferences}{0}{0}
\leavevmode\hypertarget{ref-Bolukbasi2016man}{}%
\CSLLeftMargin{{[}1{]} }
\CSLRightInline{Tolga Bolukbasi, Kai-Wei Chang, James Y. Zou, Venkatesh
Saligrama, and Adam Kalai. 2016. Man is to computer programmer as woman
is to homemaker? Debiasing word embeddings. \emph{CoRR} abs/1607.06520,
(2016). Retrieved from \url{http://arxiv.org/abs/1607.06520}}

\leavevmode\hypertarget{ref-Caliskan2017semanticsBiases}{}%
\CSLLeftMargin{{[}2{]} }
\CSLRightInline{Aylin Caliskan, Joanna J. Bryson, and Arvind Narayanan.
2017. Semantics derived automatically from language corpora contain
human-like biases. \emph{Science} 356, 6334 (April 2017), 183--186.
DOI:https://doi.org/\href{https://doi.org/10.1126/science.aal4230}{10.1126/science.aal4230}}

\leavevmode\hypertarget{ref-Garg2018years}{}%
\CSLLeftMargin{{[}3{]} }
\CSLRightInline{Nikhil Garg, Londa Schiebinger, Dan Jurafsky, and James
Zou. 2018. Word embeddings quantify 100 years of gender and ethnic
stereotypes. \emph{Proceedings of the National Academy of Sciences} 115,
16 (April 2018), E3635--E3644.
DOI:https://doi.org/\href{https://doi.org/10.1073/pnas.1720347115}{10.1073/pnas.1720347115}}

\leavevmode\hypertarget{ref-Gonen2019lipstick}{}%
\CSLLeftMargin{{[}4{]} }
\CSLRightInline{Hila Gonen and Yoav Goldberg. 2019. Lipstick on a pig:
{D}ebiasing methods cover up systematic gender biases in word embeddings
but do not remove them. In \emph{Proceedings of the 2019 conference of
the north {A}merican chapter of the association for computational
linguistics: Human language technologies, volume 1 (long and short
papers)}, Association for Computational Linguistics, Minneapolis,
Minnesota, 609--614.
DOI:https://doi.org/\href{https://doi.org/10.18653/v1/N19-1061}{10.18653/v1/N19-1061}}

\leavevmode\hypertarget{ref-Lauscher2019multidimensional}{}%
\CSLLeftMargin{{[}5{]} }
\CSLRightInline{Anne Lauscher and Goran Glavas. 2019. Are we
consistently biased? Multidimensional analysis of biases in
distributional word vectors. \emph{CoRR} abs/1904.11783, (2019).
Retrieved from \url{http://arxiv.org/abs/1904.11783}}

\leavevmode\hypertarget{ref-Manzini2019blackToCriminal}{}%
\CSLLeftMargin{{[}6{]} }
\CSLRightInline{Thomas Manzini, Yao Chong Lim, Yulia Tsvetkov, and Alan
W Black. 2019. Black is to criminal as caucasian is to police: Detecting
and removing multiclass bias in word embeddings. Retrieved from
\url{http://arxiv.org/abs/1904.04047}}

\leavevmode\hypertarget{ref-Nissim2020fair}{}%
\CSLLeftMargin{{[}7{]} }
\CSLRightInline{Malvina Nissim, Rik van Noord, and Rob van der Goot.
2020. Fair is better than sensational: Man is to doctor as woman is to
doctor. \emph{Computational Linguistics} 46, 2 (June 2020), 487--497.
DOI:https://doi.org/\href{https://doi.org/10.1162/coli_a_00379}{10.1162/coli\_a\_00379}}

\leavevmode\hypertarget{ref-Nosek2002harvesting}{}%
\CSLLeftMargin{{[}8{]} }
\CSLRightInline{Brian A. Nosek, Mahzarin R. Banaji, and Anthony G.
Greenwald. 2002. Harvesting implicit group attitudes and beliefs from a
demonstration web site. \emph{Group Dynamics: Theory, Research, and
Practice} 6, 1 (2002), 101--115.
DOI:https://doi.org/\href{https://doi.org/10.1037/1089-2699.6.1.101}{10.1037/1089-2699.6.1.101}}
